---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# mzrtsim

<!-- badges: start -->
[![Actions Status](https://github.com/yufree/mzrtsim/workflows/Render%20and%20Deploy%20RMarkdown%20Website/badge.svg)](https://github.com/yufree/mzrtsim/actions)
[![R-CMD-check](https://github.com/yufree/mzrtsim/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/yufree/mzrtsim/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The goal of mzrtsim is to make raw data and features table simulation for LC/GC-MS based  data

## Installation

You can install the development version from [GitHub](https://github.com/) with:

```{r}
# install.packages("remotes")
remotes::install_github("yufree/mzrtsim")
```
## Raw Data simulation

You could use `simmzml` to generate one mzML file.

```{r eval=FALSE}
library(mzrtsim)
data("monams1")
simmzml(db=monams1, name = 'test')
```

You will find `test.mzML` and corresponding `test.csv` with m/z, retention time and compound name of the peaks. Here the `monams1` and `monahrms1` is from the MS1 data of MassBank of North America (MoNA) and could be downloaded from their [website](https://mona.fiehnlab.ucdavis.edu/downloads). Here we only use the MS1 full scan data for simulation.

### Multiple files with experiment design

You could stimulate two groups of raw data with different peak width. Retention time could follow a uniform distribution. 100 compounds could be selected randomly and base peaks' signal to noise ratio could be sample from 100 to 1000. Each group contain 10 samples.

```{r eval=FALSE}
dir.create('case')
dir.create('control')
# 30% compounds changed between case and control
pw1 <- c(rep(5,30),rep(10,40),rep(15,30))
pw2 <- c(rep(5,20),rep(10,30),rep(15,50))
rt <- seq(10,590,length.out=100)
set.seed(1)
compound <- sample(c(1:4000),100)
set.seed(2)
sn <- sample(c(100:10000),100)
for(i in c(1:10)){
  simmzml(name=paste0('case/case',i),db=monahrms1,pwidth = pw1,compound=compound,rtime = rt, sn=sn)
}

for(i in c(1:10)){
  simmzml(name=paste0('control/control',i),db=monahrms1,pwidth = pw2,compound=compound,rtime = rt, sn=sn)
}
```

Then you could find 10 mzML files in case subfolder and another 10 mzML files in control subfolder.

### Chromatography peaks

You could also use `simmzml` to stimulate tailing/leading peaks by defining the tailing factor of the peaks. When the tailing factor is lower than 1, the peaks are leading peaks. When the tailing factor is larger than 1, the peaks are tailing peaks. 

```{r eval=FALSE}
# leading peaks
simmzml(name='test',db=monahrms1,pwidth = 10,compound=1,rtime = 100, sn=10,tailingfactor = 0.8)
# tailing peaks
simmzml(name='test',db=monahrms1,pwidth = 10,compound=1,rtime = 100, sn=10,tailingfactor = 1.5)
```

### matrix stimulation

You could also input a m/z vector as matrix mass. Those masses will generate background baseline signals. By default, the mass vector is from a previous matrix samples. 

```{r eval=FALSE}
data(mzm)
simmzml(name='test',db=monahrms1,pwidth = 10,compound=1,rtime = 100, sn=10,matrixmz = mzm,matrix = T)
```

## Batch effect simulation

You could use `mzrtsim` to make simulation of peak list.

```{r mzrtsim}
library(mzrtsim)
data("monams1")
simdata <- mzrtsim(ncomp = 100, ncond = 2, ncpeaks = 0.05,
  nbatch = 3, nbpeaks = 0.1, npercond = 10, nperbatch = c(8, 5, 7), seed = 42, batchtype = 'mb', db=monams1)
```

Here we make a simulation of 100 compounds from selected database with two conditions and three batches. 5 percentage of the peaks were influnced by conditions and 10 percentage of the peaks were influnced by batch effects. Three different type could be simulated: monotonic, random and block. You could also bind batchtype, for example, 'mb' means the simulation would contain both monotonic and block batch effects. 'db' means the spectra database to be used for simulation. 'monams1' means LC-MS spectra from MoNA amd 'hmdbcms' means GC-EI-MS spectra from HMDB.

You could save the simulated data into multiple csv files by `simdata` function. `simraw.csv` could be used for metaboanalyst. `simraw2.csv` show the raw peaks list. `simcon.csv` show peaks influnced by conditions only.`simbatchmatrix.csv` show peaks influnced by batch effects only. `simbat.csv` show peaks influnced by batch effects and conditions. `simcomp.csv` show independent peaks influnced by conditions and batch effects. `simcompchange.csv` show the conditions changes of each groups. `simblockbatchange.csv` show the block batch changes of each groups. `simmonobatchange.csv` show the monotonic batch changes of each groups.

```{r simdata, eval=F}
simdata(sim,name = "sim")
```

## Basic models of batch effects correction

Batch effects are the variances caused by factor other than the experimental design. We could simply make a linear model for the intensity of one peak:

$$Intensity =  Average + Condition + Batch + Error$$

Research is focused on condition contribution part and overall average or random error could be estimated. However, we know little about the batch contribution. Sometimes we could use known variables such as injection order or operators as the batch part. However, in most cases we such variable is unknown. Almost all the batch correction methods are trying to use some estimations to balance or remove the batch effect.

For analytical chemistry, internal standards or pool quality control samples are actually standing for the batch contribution part in the model. However, it's impractical to get all the internal standards when the data is collected untargeted. For methods using internal standards or pool quality control samples, the variations among those samples are usually removed as median, quantile, mean or the ratios. Other ways like quantile regression, centering and scaling based on distribution within samples could be treated as using the stable distribution of peaks intensity to remove batch effects. However, here we would turn to the experiments without pooled quality control samples and correction methods which make estimation of batch effects. Some combination methods using pooled quality control samples  or control samples to estimate the batch effect like Remove Unwanted Variation(RUV) would not be covered here.

### Distribution of intensity

Intensity collects from LC/GC-MS always showed a right-skewed distribution. Log transformation is often necessary for further statistical analysis. In some case, a Log-transformed intensity could be visualized easily.

## Evaluation of batch correction

Various methods have been used for batch correction and evaluation. For our simulation, we need some methods to check the performances of correction. Difference analysis would be a common method for evaluation. Then we could check whether this peak is true positive or false positive by settings of the simulation. 

