---
title: "mzrtsim: LC/GC-MS Raw Data and Peak List Simulation"
author:
  - name: Miao YU
    affiliation: Department of Environmental Medicine and Climate Science, Icahn School of Medicine at Mount Sinai
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{mzrtsim: LC/GC-MS Raw Data and Peak List Simulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

`mzrtsim` generates simulated LC/GC-MS data at two levels:

1. **Raw data simulation** — produces `.mzML` files from real spectral
   databases (MoNA, HMDB), with realistic chromatographic peak shapes,
   tailing, noise, and matrix background.
2. **Peak list simulation** — produces feature tables with configurable
   condition effects and batch effects for benchmarking downstream
   statistical workflows.

# Installation

```{r install, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("mzrtsim")
```

# Raw Data Simulation

## Basic Usage

`simmzml()` generates one `.mzML` file and a companion `.csv` file
containing ground-truth peak information (m/z, retention time, database
intensity, simulated maximum intensity, compound name).

```{r basic-sim, eval=FALSE}
library(mzrtsim)
data("monahrms1")
simmzml(db = monahrms1, name = "test", n = 5,
        rtrange = c(0, 60), scanrate = 0.5, seed = 42)
```

The output CSV now includes a `sim_ins` column — the absolute
ground-truth maximum intensity of each simulated peak, accounting for
response factor, peak height, and chromatographic profile. This is useful
for evaluating peak detection algorithms.

## Tailing and Leading Peaks

Control chromatographic peak shape via `tailingfactor`:

```{r tailing, eval=FALSE}
# Tailing peaks (tailingfactor > 1)
simmzml(db = monahrms1, name = "tailing", n = 3,
        pwidth = 10, tailingfactor = 1.5, seed = 42)

# Leading peaks (tailingfactor < 1)
simmzml(db = monahrms1, name = "leading", n = 3,
        pwidth = 10, tailingfactor = 0.8, seed = 42)
```

## Gradient Baseline Simulation

The `baseline` parameter now accepts a numeric vector to simulate
gradient elution baseline drift:
```{r gradient, eval=FALSE}
rtime0 <- seq(0, 600, 0.2)
dynamic_baseline <- seq(100, 500, length.out = length(rtime0))
simmzml(db = monahrms1, name = "gradient", n = 5,
        baseline = dynamic_baseline, seed = 42)
```

## Blank / Matrix Simulation

`simmzml_blank()` generates mzML files containing only background noise
and optional matrix peaks — useful for generating blank controls or
testing baseline correction algorithms:

```{r blank, eval=FALSE}
# Pure noise blank
simmzml_blank("pure_blank", rtrange = c(0, 60))

# Blank with serum matrix peaks
data(mzm)
simmzml_blank("serum_blank", matrixmz = mzm)
```

## Background Injection

You can inject simulated peaks into a real blank mzML file using the
`background` parameter in `simmzml()`:

```{r background, eval=FALSE}
simmzml(db = monahrms1, name = "injected", n = 10,
        background = "real_blank.mzML", seed = 42)
```

# Peak List Simulation

`mzrtsim()` generates feature tables with controlled condition and batch
effects for benchmarking normalisation and batch correction methods.

```{r peaklist}
library(mzrtsim)
data("hmdbcms")
sim <- mzrtsim(ncomp = 20, ncond = 2, ncpeaks = 0.1,
               nbatch = 2, nbpeaks = 0.1, npercond = 5,
               nperbatch = c(5, 5), seed = 42,
               batchtype = "mb", db = hmdbcms)
str(sim, max.level = 1)
```

The returned list includes the data matrix, group assignments, and
separate matrices for condition-only and batch-only effects. Use
`simdata()` to export all components as CSV files.

# Batch Correction Utilities

The package provides several QC-based batch correction methods:

```{r bc}
peak <- c(100, 200, 300, 400, 500)
qc <- c(150, 250, 350)

# Centering
bccenter(peak, qc, log = TRUE)

# Pareto scaling
bcpareto(peak, qc, log = FALSE)
```

Available methods: `bccenter`, `bcscaling`, `bcpareto`, `bcrange`,
`bcvast`, `bclevel`.

# Session Info

```{r sessioninfo}
sessionInfo()
```
